Mixture Models for Single Cell Assays

### 2012-03-12 MIMOSA.tex
gradient and hessian for responder and non-responder marginal likelihoods
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

and the Hessian is:
\begin{align}
j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{iuj})
\end{align}

The marginal likelihood of the responding component in eq~\ref{eq:postmult} can be rewritten as:
\begin{align}
\begin{split}
\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
\end{split}
\end{align}
with gradient:
\begin{align}
\frac{\partial l}{\partial\alpha_{\phi j}} = \sum_i \Psi(\alpha_{\phi j}+n_{\phi ij})-\Psi(\alpha_{\phi j})+\Psi(\sum_j \alpha_{\phi j}) - \Psi(\sum_j \alpha_{\phi j}+n_{\phi ij}): (\phi \in {s,u})
\end{align}
and Hessian:
\begin{align}
\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
\frac{\partial^2l}{\partial \alpha_{\phi j}^2}&=\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
\end{align}

>>>>

and the Hessian is:
\begin{align}
j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{uj})
\end{align}

The marginal likelihood of the responding component in eq~\ref{eq:postmult} can be rewritten as:
\begin{align}
\begin{split}
\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
\end{split}
\end{align}
with gradient:
\begin{align}
\frac{\partial l}{\partial\alpha_{\phi j}} = \sum_i \Psi(\alpha_{\phi j}+n_{\phi ij})-\Psi(\alpha_{\phi j})+\Psi(\sum_j \alpha_{\phi j}) - \Psi(\sum_j \alpha_{\phi j}+n_{\phi ij}): (\phi \in {s,u})
\end{align}
and Hessian:
\begin{align}
\frac{\partial^2l}{\partial \alpha_{\phi j}^2}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}+n_{\phi ij}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}&\overset{k\ne j}{=}\sum_i\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
\end{align}
